Model for DES Type Ia Supernova Cosmology Analysis

### 2015-09-19 doc/model.tex
Nice happy version of things

The second term gives the type probability. In our model,
the underlying relative rates between SNe~Ia and non-Ia is linear in redshift,
 parameterized by the
its value at $z=0$ and $1.1 \times z_{max}$, 
\begin{equation}
\hat{\theta}_{r}=\theta_{r0}+z\left(\frac{\theta_{r1}-\theta_{r0}}{1.1 \times z_{max}}\right).
\end{equation}
The relative rate for SN~Ia discovery for the truncated distribution is
\begin{equation}
\theta_{r}=\frac{\hat{\theta}_{r}}{\hat{\theta}_{r}P_{Ia}(\mathit{ADU} > \mathit{ADU}_0) + (1-\hat{\theta}_{r})P_{non-Ia}(\mathit{ADU} > \mathit{ADU}_0)}.
\end{equation}
The spectroscopic classifications are drawn from
\begin{equation}
T_S | \Omega_M, w, \theta_{r0}, \theta_{r1} , \alpha_{Ia},\sigma_{Ia}, \alpha_{\mathit{non-Ia}},\sigma_{\mathit{non-Ia}} \sim \text{Bernoulli}(\theta_r).
\end{equation}
Although the underlying type distribution depends only on $ \theta_{r0}$ and $ \theta_{r1}$,  
the observed truncated distribution does depend on the other parameters through 
$P_{X}(\mathit{ADU} > \mathit{ADU}_0)$.

@@

The results that follow are based on the parameter
choices:
\begin{itemize}
\item Cosmological parameters $\Omega_M=0.28$, $w=-1$.
\item Type~Ia supernovae with intrinsic dispersion $\sigma_{SNIa}=0.1$.
\item Non-Ia supernovae 2 magnitudes fainter than SNe~Ia with intrinsic
dispersion $\sigma_{non-Ia}=1$.
\item $\theta_r=r_0 + (r_1-r_0)z/1.54$ for a constant relative fraction of SNe~Ia where $r_1=0.2$,
$r_0=0.95$.
\item 2000 transients distributed by co-moving volume in the range $0.1<z<1.4$.
\item Detection threshold at approximately the mean brightness SN~Ia at $z=1.2$.
\item $z_{min}=0.1/1.1$ and $z_{max}=1.4\times 1.1$ is the range of possible misassigned
host redshift.
\item $P_{host}=0.98$ for the probability of correct host-galaxy assignment.
\item Accumulated spectra for 20\%, 60\%, and 100\% of discovered transients.
\end{itemize}

A prior is applied to the SN~Ia absolute magnitude and intrinsic dispersion
\begin{align}
\alpha_{SNIa} & \sim \mathcal{N}_{\ln}\left(\ln\left(\alpha^0_{SNIa}\right),\frac{\ln{10}}{2.5}0.02\right)\\
\sigma_{SNIa} & \sim \mathcal{N}_{\ln}\left(\ln\left(0.1\right),0.1\right),
\end{align}
where $\alpha^0_{SNIa}$ is the true input luminosity.

The simulated data are analyzed using 
\citet{stan-software:2015}, a
a probabilistic programming language for
inference of Bayesian models with Hamiltonian Monte Carlo.

\subsubsection{Results}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[scale=0.35]{/Users/akim/project/abc/results/temp/{contour..2000..noMalm.1.0}.pdf} \newline
\caption{Probability distribution function for $\Omega_M$--$w$,
 100\% spectroscopy no Malmquist bias.
   Solid lines represent the input cosmology.  Dashed lines in the
   histograms are the 0.16 and 0.86 quantiles. 
   \label{contour1:fig}}
\end{figure}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[scale=0.35]{/Users/akim/project/abc/results/temp/{contour..2000.1.0}.pdf} \newline
\caption{Probability distribution function for $\Omega_M$--$w$,
 100\% spectroscopy with Malmquist bias.
   \label{contour2:fig}}
\end{figure}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[scale=0.35]{/Users/akim/project/abc/results/temp/{contour..2000.0.6}.pdf} \newline
\caption{Probability distribution function for $\Omega_M$--$w$,
 60\% spectroscopy with Malmquist bias.
   \label{contour3:fig}}
\end{figure}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[scale=0.35]{/Users/akim/project/abc/results/temp/{contour..2000.0.2}.pdf} \newline
\caption{Probability distribution function for $\Omega_M$--$w$,
 20\% spectroscopy with Malmquist bias.
   \label{contour4:fig}}
\end{figure}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[scale=0.35]{/Users/akim/project/abc/results/temp/{contour..2000..noMalm.0.2}.pdf} \newline
\caption{Probability distribution function for $\Omega_M$--$w$,
 20\% spectroscopy without Malmquist bias.
   \label{contour4:fig}}
\end{figure}

\begin{table}
\centering
\begin{tabular}{cc|c|cc|}
\hline
ADU0&\% Spec&\%-ile & $w$ range & $\Delta w$\\ \hline
0 & 100 &$0.68$ & $[-1.174, -0.982]$ & $0.192$ \\
&&$0.90$ & $[-1.239, -0.923]$ & $0.316$ \\
&&$0.95$ & $[-1.267, -0.894]$ & $0.373$ \\ \hline
0.2 & 100&$0.68$ & $[-1.156, -0.972]$ & $0.184$ \\
&&$0.90$ & $[-1.222, -0.912]$ & $0.309$ \\
&&$0.95$ & $[-1.255, -0.883]$ & $0.372$ \\ \hline
0.2&60&$0.68$ & $[-1.209, -1.014]$ & $0.195$ \\
&&$0.90$ & $[-1.275, -0.955]$ & $0.321$ \\
&&$0.95$ & $[-1.309, -0.924]$ & $0.385$ \\ \hline
0&20&$0.68$ & $[-1.226, -1.032]$ & $0.194$ \\
&&$0.90$ & $[-1.289, -0.969]$ & $0.320$ \\
&&$0.95$ & $[-1.322, -0.942]$ & $0.379$ \\ \hline
0.2& 20&$0.68$ & $[-1.229, -1.033]$ & $0.196$ \\
&&$0.90$ & $[-1.294, -0.972]$ & $0.322$ \\
&&$0.95$ & $[-1.323, -0.943]$ & $0.380$ \\
\hline
\end{tabular}
\caption{Credible intervals of $w$ using the SNe~Ia only with spectroscopic
observations.\label{ia_only:tab}}
\end{table}

>>>>

The second term gives the type probability. In our model,
the underlying relative rates between SNe~Ia and non-Ia is linear in redshift,
 parameterized by the
its value at $z=0$ and $1.1 \times z_{max}$, 
\begin{equation}
\hat{\theta}_{r}=\theta_{r0}+z\left(\frac{\theta_{r1}-\theta_{r0}}{1.1 \times z_{max}}\right).
\end{equation}
The relative rate for SN~Ia discovery for the truncated distribution is
\begin{equation}
\theta_{r}=\frac{\hat{\theta}_{r}P_{Ia}(\mathit{ADU} > \mathit{ADU}_0)}{\hat{\theta}_{r}P_{Ia}(\mathit{ADU} > \mathit{ADU}_0) + (1-\hat{\theta}_{r})P_{non-Ia}(\mathit{ADU} > \mathit{ADU}_0)}.
\end{equation}
The spectroscopic classifications are drawn from
\begin{equation}
T_S | \Omega_M, w, \theta_{r0}, \theta_{r1} , \alpha_{Ia},\sigma_{Ia}, \alpha_{\mathit{non-Ia}},\sigma_{\mathit{non-Ia}} \sim \text{Bernoulli}(\theta_r).
\end{equation}
Although the underlying type distribution depends only on $ \theta_{r0}$ and $ \theta_{r1}$,  
the observed truncated distribution does depend on the other parameters through 
$P_{X}(\mathit{ADU} > \mathit{ADU}_0)$.

@@

The results that follow are based on the parameter
choices:
\begin{itemize}
\item Cosmological parameters $\Omega_M=0.28$, $w=-1$.
\item Type~Ia supernovae with intrinsic dispersion $\sigma_{SNIa}=0.1$.
\item Non-Ia supernovae 2 magnitudes fainter than SNe~Ia with intrinsic
dispersion $\sigma_{non-Ia}=1$.
\item $\theta_r=r_0 + (r_1-r_0)z/1.54$ for a constant relative fraction of SNe~Ia where $r_1=0.2$,
$r_0=0.95$.
\item 2000 transients distributed by co-moving volume in the range $0.1<z<1.4$.
\item Detection threshold at approximately the mean brightness SN~Ia at $z=1$.
\item $z_{min}=0.1/1.1$ and $z_{max}=1.4\times 1.1$ is the range of possible misassigned
host redshift.
\item $P_{host}=0.98$ for the probability of correct host-galaxy assignment.
\item Accumulated spectra for 0\%, 20\%, 60\%, and 100\% of discovered transients.
\end{itemize}
This setup is referred to as Model 1.

The Hubble diagram for a realization drawn from this model is shown in Figure~\ref{hd:fig}.
Type~Ia's are shown in blue and non-Ia's in red, those that fall below detection threshold
are faded.  The solid points
represent those objects spectroscopically typed for the case of 20\% follow-up.
The green X's represent the 2\% among those without spectroscopy that
are assigned an incorrect redshift.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[scale=0.4]{/Users/akim/project/abc/doc/seed2.pdf}
   \includegraphics[scale=0.4]{/Users/akim/project/abc/doc/seed2_pop2.pdf}  
\caption{Hubble diagrams for a realization drawn from the baseline Model 1 (left)
and Model 2 with a second non-Ia population (right).
Type~Ia's are shown in blue and non-Ia's in red, those that fall below detection threshold
are faded.  The green X's represent the 2\% among those without spectroscopy that
are assigned an incorrect redshift.
   \label{hd:fig}}
\end{figure}

The simulated data are analyzed using 
\citet{stan-software:2015},
a probabilistic programming language for
inference of Bayesian models with Hamiltonian Monte Carlo.
In addition to the simulated data,
supplemental priors are applied to the SN~Ia absolute magnitude and intrinsic dispersion
\begin{align}
\alpha_{SNIa} & \sim \mathcal{N}_{\ln}\left(\ln\left(\alpha^0_{SNIa}\right),\frac{\ln{10}}{2.5}0.02\right)\\
\sigma_{SNIa} & \sim \mathcal{N}_{\ln}\left(\ln\left(0.1\right),0.1\right),
\end{align}
where $\alpha^0_{SNIa}$ is the true input luminosity.  This prior is added in lieu of
a pure large SN~Ia sample at low-redshift.

The results of the analyses meet expectations.  As the input data and analyses are
based on the same model, the input parameters lie in reasonable position
within the resultant posterior despite the non-trivial appearance of
the Hubble diagram in Figure~\ref{hd:fig}.
The parameter pdf's are shown in Figure~\ref{contour1:fig} for the ideal case of 100\% spectroscopic classification.  The input values for all 7 parameters fall within the 68\%
credible interval.  The same is true for the unideal case of no
spectroscopic classification.  Table~\ref{seed2:tab} lists the modes
of the $w$ distribution for varying spectroscopic coverage: there is no appreciable
bias relative to the statistical uncertainties.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[scale=0.17]{/Users/akim/project/abc/results/seed2/{contour.2000.1.0}.pdf}
   \includegraphics[scale=0.17]{/Users/akim/project/abc/results/seed2_pop2/{contour.2000.1.0}.pdf} 
\caption{Probability distribution functions of the model parameters for the case of
 100\% spectroscopy for Model 1 (left) and Model 2 (right).
    Solid lines represent the input parameters.  Dashed lines in the
   histograms are the 0.16 and 0.84 quantiles. 
   \label{contour1:fig}}
\end{figure}

\begin{table}
\centering
\begin{tabular}{|c|c|ccc|}
\hline
Fraction Spec&\%-ile & $w$ &$w$ range & $\Delta w$\\ \hline
$1.00$ &$0.68$ &$-1.08$ & $[-1.196, -0.914]$ & $0.281$ \\
$1.00$ &$0.90$ &$-1.08$ & $[-1.292, -0.826]$ & $0.466$ \\
$1.00$ &$0.95$ &$-1.08$ & $[-1.340, -0.784]$ & $0.556$ \\
$0.60$ &$0.68$ &$-1.08$ & $[-1.227, -0.933]$ & $0.293$ \\
$0.60$ &$0.90$ &$-1.08$ & $[-1.331, -0.836]$ & $0.495$ \\
$0.60$ &$0.95$ &$-1.08$ & $[-1.385, -0.787]$ & $0.598$ \\
$0.20$ &$0.68$ &$-1.15$ & $[-1.249, -0.943]$ & $0.305$ \\
$0.20$ &$0.90$ &$-1.15$ & $[-1.354, -0.851]$ & $0.502$ \\
$0.20$ &$0.95$ &$-1.15$ & $[-1.408, -0.810]$ & $0.597$ \\
$0.00$ &$0.68$ &$-1.09$ & $[-1.218, -0.909]$ & $0.309$ \\
$0.00$ &$0.90$ &$-1.09$ & $[-1.320, -0.812]$ & $0.508$ \\
$0.00$ &$0.95$ &$-1.09$ & $[-1.372, -0.770]$ & $0.602$ \\
\hline
\end{tabular}
\caption{Mode and credible intervals of $w$ for Model 1 for varying
fractions of spectroscopic coverage. \label{seed2:tab}}
\end{table}

%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
%   \centering
%   \includegraphics[scale=0.17]{/Users/akim/project/abc/results/seed2/{contour.2000.0.0}.pdf}
%   \includegraphics[scale=0.17]{/Users/akim/project/abc/results/seed2_pop2/{contour.2000.0.0}.pdf} 
%\caption{Probability distribution functions for the difference in parameters
%with 100\% and 0\% spectroscopic folowup for Model 1 (left) and Model 2 (right).  \label{contour2:fig}}
%\end{figure}

Though the decrease in spectroscopic coverage produces low bias, it does lead to
increased parameter uncertainties.
The credible intervals for $w$ are shown in Table~\ref{seed2:tab}:
for the case of 100\% follow-up the range of the 68\% $w$ interval is $\Delta w = 0.281$. 
At the opposite extreme of no spectroscopic classification the $w$ interval inflates to
$\Delta w = 0.309$, a constraining result nevertheless considering that none of the
objects are directly typed and redshifted.

As a stress test, we generate a simulated data based on a different model
referred to as Model 2, which
has a second non-Ia population with average luminosity 0.5 mag fainter
than the Ia's and 0.25 mag dispersion.
The relative rates of the original non-Ia population relative to the second is quadratic:
\begin{equation}
1 -\frac{0.8}{1.54}z^2.
\end{equation}
A realization of transients from this model, using the same random seeds
as Model 1, is shown in Figure~\ref{hd:fig}.

When there is 100\% spectroscopic coverage the   $\Omega_M$--$w$  posterior accommodates the input 
parameters, as shown  in
Figure~\ref{contour1:fig}.
A salient feature that contrasts with Model 1 is 
seen in the modes for $w$ listed in Table~\ref{seed2_pop2:tab}: biases
larger than the statistical uncertainties appear with decreasing
spectroscopic coverage.  Note that the Hubble diagrams
in these analyses are almost identical: there are redshift misassignments
for 2\% of the transients that lack spectroscopy.

%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
%   \centering
%   \includegraphics[scale=0.35]{/Users/akim/project/abc/results/seed2_pop2/{contour.2000.1.0}.pdf} \newline
%\caption{Probability distribution function for $\Omega_M$--$w$,
% 100\% spectroscopy, biased.
%    Solid lines represent the input cosmology.  Dashed lines in the
%   histograms are the 0.16 and 0.84 quantiles. 
%   \label{contour3:fig}}
%\end{figure}
%
%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
%   \centering
%   \includegraphics[scale=0.35]{/Users/akim/project/abc/results/seed2_pop2/{contour.2000.0.0}.pdf} \newline
%\caption{Probability distribution function for $\Omega_M$--$w$,
% no spectroscopy, biased.
%   \label{contour4:fig}}
%\end{figure}

\begin{table}
\centering
\begin{tabular}{|c|c|ccc|}
\hline
Fraction Spec&\%-ile &$w$ & $w$ range & $\Delta w$\\ \hline
$1.00$ &$0.68$ &$-0.94$ & $[-1.078, -0.804]$ & $0.274$ \\
$1.00$ &$0.90$ &$-0.94$ & $[-1.168, -0.720]$ & $0.447$ \\
$1.00$ &$0.95$ &$-0.94$ & $[-1.214, -0.680]$ & $0.534$ \\
$0.60$ &$0.68$ &$-1.09$ & $[-1.182, -0.883]$ & $0.298$ \\
$0.60$ &$0.90$ &$-1.09$ & $[-1.285, -0.791]$ & $0.494$ \\
$0.60$ &$0.95$ &$-1.09$ & $[-1.332, -0.749]$ & $0.583$ \\
$0.20$ &$0.68$ &$-1.09$ & $[-1.247, -0.936]$ & $0.311$ \\
$0.20$ &$0.90$ &$-1.09$ & $[-1.359, -0.840]$ & $0.518$ \\
$0.20$ &$0.95$ &$-1.09$ & $[-1.413, -0.798]$ & $0.614$ \\
$0.00$ &$0.68$ &$-1.11$ & $[-1.266, -0.948]$ & $0.317$ \\
$0.00$ &$0.90$ &$-1.11$ & $[-1.375, -0.850]$ & $0.525$ \\
$0.00$ &$0.95$ &$-1.11$ & $[-1.428, -0.808]$ & $0.621$ \\
\hline
\end{tabular}
\caption{Mode and credible intervals of $w$ for Model 2  for varying
fractions of spectroscopic coverage.\label{seed2_pop2:tab}}
\end{table}

Some initial conclusions.
\begin{itemize}
\item Knowing the underlying model that produces data, you should
get results statistically consistent with the truth.  A tautology?
\item Typing is important when the non-Ia model is not accurate.
\end{itemize}
